\textit{MSF: Modulated Sub-graph Finder}

### 2018-03-20 MSF_Paper_FA.tex
MSF_Paper_FA.tex

\TODO{FA until here}.

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach in Java to
find the modulated sub-graphs of gene interactions from the cell
signaling network. \texttt{MSF} does not use predefined gene sets in
functional units, it builds up the sub-graphs from scratch using the
whole cell signaling network from Reactome. The network has nodes
corresponding to genes and edges representing their directed
regulating interactions.

\texttt{MSF} uses the individual gene's \textit{p}-values generated
from the DGEA. The \textit{p}-value expresses the probability that the
null hypothesis of unmodified gene expression can be rejected for a
given statistical model. To find significantly modulated sub-graphs
individual \textit{p}-values of the vicinal genes in the global
network are combined into a single combined \textit{p}-value, using a
statistical method for combining dependent \textit{p}-values described
by Hartung~\cite{Hartung}. Hartung uses the inverse of standard normal
distribution function, individual gene \textit{p}-values are first
transformed to their corresponding normal score. Then using these
normal scores, the correlation between genes is calculated, the normal
scores and correlation are applied to the inverse normal function to
calculate the individual \textit{p}-value for a sub-graph. An overall
\textit{p}-value of a sub-graph will express the significance of all
genes in the sub-graph being modulated together.

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a three steps heuristic as described in the
following.The pipeline of identification of modulated sub-graphs from
a network by \texttt{MSF} is depicted in Fig (1).
\newline

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value from
the network. It tries to extend the sub-graph to the most significant
neighboring gene. A single combined \textit{p}-value is calculated for
the two genes. If the combined \textit{p}-value is smaller than the
minimal individual gene's \textit{p}-value, the extended sub-graph is
accepted. Then the next most significant neighboring gene is added to
the sub-graph and the combined \textit{p}-value is calculated, if the
new combined \textit{p}-value of three genes is smaller than the
combined \textit{p}-value of the first two genes, the extended
sub-graph is accepted. This step is iteratively repeated until no
further extension is accepted. In this case the process starts over
with all remaining genes not yet in a significantly modulated
sub-graph. This step identifies all the trivial sub-graphs that are
modulated in the whole network~(Fig.~1b).\newline

@@

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood.
This is done by testing all possible extension by N genes (default two
genes) for all genes in the sub-graph. Again, this step is iteratively
repeated until no further genes are added to the significant
differentially expressed sub-graphs. This steps bridges small gaps of
genes without a clear differential signal in the
DGEA~(Fig.~1c).\newline

@@

After detection and extension of the modulated sub-graphs, they are
tested if combined sub-graph score is better than on their own. The
merging of the two sub-graphs is done by depth first search. If the
two sub-graphs merge with the connector of at most N genes (default
two genes) and the combined \textit{p}-value of the merged sub-graph
including the bridging genes in between is less than the individual
\textit{p}-values of the two sub-graphs, the two sub-graphs are merged
together to one big modulated sub-graph~(Fig.~1d). This step is
repeated iteratively until no sub-graphs could be merged.\newline

@@

In a last post processing step \texttt{MSF} identifies the trigger
points of the modulated sub-graphs. These trigger genes are the
sources of the sub-graphs with only outgoing edges. These genes can be
interpreted as the possible entry points of perturbation from where
the stimulus causes downstream effects. In the same spirit the most
downstream genes of the modulated sub-graph are identified and defined
as sinks. Sinks can be interpreted as the effectors where the
integrated information within the signal transduction network is set
to action. Genes in circular loops in the sub-graphs have no source or
sink.

@@

To assess the robustness and stability of our method, Poisson
distributed noise was added to the sample read counts. Then DGEA was
carried out on the disturbed data with the same parameters as for the
native data, followed by analysis with \texttt{MSF}. Using this
criteria, the goal was to spot almost identical modulated sub-graphs
in the real and noisy data. This procedure was carried out 100 times
and every time the genes from the sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the
native data. This was also done for the DEG obtained for each run, a
cutoff was used to compare the retrieved DEG from noisy data to the
native data.  The robustness of \texttt{MSF} and the DEG analysis for
the three time-points are shown in~(Fig.~2). We were very stringent in
adding noise to the counts, which could be seen by the DEG analysis
robustness. For the last time-point (48hpi) 84\% genes of the
sub-graphs from native data were retrieved, for 24hpi robustness was
about 79\%. Robustness was 71\% at 6hpi, this could be due to that the
infection merely started and the signal transdcution was weak as
compared to the later time-points.

Simulation testing was also conducted on \texttt{MSF}, the objective
was to find less sources than observed in the sub-graphs identified by
\texttt{MSF} from native data. Simulation results showed a chance of
less than 10 percent to get more sources than observed.

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. The aim of
testing the EBOV infection data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources.

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
list generated by edgeR was used as an input for \texttt{MSF}. The
EBOV infection data has three time-points 6, 24 and 48 hour post
infection (hpi). At 6hpi, five modulated sub-graphs were
identified. Most of the genes part of the sub-graphs were chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24hpi nine
modulated sub-graphs were identified, IFNA1 was identified as one of
the sources. For the last time-point again IFNB1 and IFNA1 were
identified as the two sources out of the possible sources from seven
modulated sub-graphs identified.

@@

\subsection*{Modulated sub-graphs at 6hpi}

 Three main modulated sub-graphs identified by \texttt{MSF} at 6hpi
 are shown in~(Fig.~3). The graph shows different colored nodes as
 genes associated to KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. The graph
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

At 6hpi Cytokegg identified the sub-graph genes to be enriched in
Toll-like receptor signaling, TNF signaling, IL-17 signaling and
NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The
earliest response EBOV induces on cytokine is via TLR4-mediated
signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes
at 6hpi showed toll-like receptor signaling being most significantly
dis-regulated pathway, this important signaling pathway was not
identified by SPIA along with RIG-I-like receptor signaling pathway,
TNF signaling pathway and IL-17 signaling (Table 1 in the supplement
material). On the later time-points \texttt{MSF} and SPIA showed
consensus on the important pathways e.g Chemokine signaling,
Cytokine-Cytokine receptor interaction, NF-kappa B signaling,
Cytosolic DNA-sensing and RIG-I-like receptor signaling as being
dis-regulated (Table 2-3 in the supplement).

@@

\begin{figure*}[ht!]
		\centering
	\fbox{
		\begin{minipage}{17 cm}

	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{Tgfb1.png}
		\caption{Input}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{TGFBstep1.png}
		\caption{Step 1}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{TGFBstep2.png}
		\caption{Step 2}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=10.0cm]{TGFBstep3.png}
		\caption{Step 3}
	\end{subfigure}
	\caption{Overview of \texttt{MSF} process showing 3 step toy
          example to identify the modulated sub-graphs. (a) Input:
          showing a network of genes with green nodes as genes with
          \textit{p}-values $< $ 0.05, and blue nodes are genes with
          \textit{p}-values $>$ 0.05, (b) Step 1: genes circled red
          show the two initial modulated sub-graphs identified,
          sub-graph 1 with genes M,K,C and sub-graph 2 with genes
          S,V,T, (c) Step 2: shows the modulated sub-graph 1 being
          extended to M,K,C,A,D while no extension in sub-graph 2 (d)
          Step 3: shows both the modulated sub-graphs merging together
          with the addition of 3 genes Z,W,P.}
			\end{minipage}
}
\end{figure*}

@@

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
	\includegraphics[width=1.05\linewidth]{SeperatedMergedNetwork.png}
	\caption{Shows the three modulated sub-graphs identified by
          MSF at 6hpi, where nodes are genes and edges are the
          interactions. The node coloring is associated to KEGG
          pathways referring to the colors in the legend. The graph
          edges are from Reactome. Important genes to EBOV infection
          as from literature are enlarged in the graph. The signal
          flow corresponding to each sub-graph is depicted as simple
          flow chart against each sub-graph.}
		\end{minipage}
}
\end{figure*}

>>>>

\section*{Methods}
\texttt{MSF} is developed as a novel heuristic approach to
find concertedly modulated sub-graphs in networks of biological interactions.
\texttt{MSF} does not use predefined gene sets grouped into
functional units, but rather relies purely on the network of interacting genes.
The inputed network consists of nodes corresponding to genes and edges
representing interactions. Furthermore it utilizes comprehensive results from a
differential gene expression analysis to discover the sub-graphs, or modules, which
are as a whole modulated.

\texttt{MSF} uses the individual gene's \textit{p}-values generated
from the DGEA. The \textit{p}-value expresses the probability that the
null hypothesis of unmodified gene expression can be rejected for a
given statistical model. To find significantly modulated sub-graphs
individual \textit{p}-values of the vicinal genes in the global
network are combined into a single combined \textit{p}-value, using a
statistical method for combining dependent \textit{p}-values described
by Hartung~\cite{Hartung}. Hartung uses the inverse of standard normal
distribution function, individual gene \textit{p}-values are first
transformed to their corresponding normal score. Then using these
normal scores, the correlation between genes is calculated, the normal
scores and correlation are applied to the inverse normal function to
calculate the combined \textit{p}-value for all genes examined, namely the
examined sub-graph. The combined \textit{p}-value of a sub-graph will
express the significance of all genes in the sub-graph being modulated
together. Thereby, the information from the different genes are used as,
although not independent, replicated measurement of the behavior of the whole
sub-graph. This potentially increases the power to detect also significant
sub-graphs consisting of genes which are not significant on there own.
\newline

@@

To reduce the complexity to score all possible connected sub-graphs
\texttt{MSF} applies a three steps heuristic as described in the
following. The proceeding identification of modulated sub-graphs from
a network by \texttt{MSF} is presented graphical and as pseudocode in
Fig.~\ref{fig:algorithm}. \TODO{add pseudocode or better replace figure with pseudocode}

\begin{figure}[ht!]
		\centering
	\fbox{
		\begin{minipage}{8 cm}

	\begin{subfigure}{.6\textwidth}
		\centering
		\includegraphics[width=6.5cm]{Tgfb1.png}
		\caption{Input}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep1.png}
		\caption{Step 1}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep2.png}
		\caption{Step 2}
	\end{subfigure}
	\begin{subfigure}{.5\textwidth}
		\centering
		\includegraphics[width=6.5cm]{TGFBstep3.png}
		\caption{Step 3}
	\end{subfigure}
	\caption{Overview of \texttt{MSF} process steps exemplified on a simplified toy
          example. (a) Input:
          showing a network of genes with green nodes as genes with
          \textit{p}-values $< $ 0.05, and blue nodes are genes with
          \textit{p}-values $>$ 0.05, (b) Step 1: genes circled red
          show the two initial modulated sub-graphs identified,
          sub-graph 1 with genes M,K,C and sub-graph 2 with genes
          S,V,T, (c) Step 2: shows the modulated sub-graph 1 being
          extended to M,K,C,A,D while no extension in sub-graph 2 (d)
          Step 3: shows both the modulated sub-graphs merging together
          with the addition of 3 genes Z,W,P.\label{fig:algorithm}}
			\end{minipage}
}

\end{figure}

@@

\texttt{MSF} constructs the first sub-graph starting with the genes
associated with the lowest (most significant) \textit{p}-value deduced from
the DGEA. From this seed the sub-graph is attempted to be extended with its
neighboring genes, starting with the most significant one. A single combined
\textit{p}-value is calculated for
the two genes. If the combined \textit{p}-value is smaller than the
minimal individual gene's \textit{p}-value, the extended sub-graph is
accepted. This step is iteratively repeated until no
further extension is accepted. In this case the process starts over
with all remaining genes not yet in a significantly modulated
sub-graph. This step identifies all the trivial sub-graphs that are
modulated in the whole network (Fig.~\ref{fig:algorithm}b).\newline

@@

In the next step initial modulated sub-graphs are used to check if
they could further be extended beyond the immediate neighborhood.
This is done by testing all possible extension paths up to N genes
for all genes in the sub-graph. Again, this step is iteratively
repeated until no further genes are added to the significant
differentially expressed sub-graphs. This steps bridges small gaps of
genes without a clear differential signal in the
DGEA~(Fig.~\ref{fig:algorithm}c).\newline

@@

After detection and extension of the modulated sub-graphs, they are
tested if combined sub-graph score is better than on their own. The
merging of the two sub-graphs is done by depth first search. If the
two sub-graphs merge with the connector of at most N genes (default
two genes) and the combined \textit{p}-value of the merged sub-graph
including the bridging genes in between is less than the individual
\textit{p}-values of the two sub-graphs, the two sub-graphs are merged
together to one big modulated sub-graph~(Fig.~\ref{fig:algorithm}d). This step is
repeated iteratively until no sub-graphs could be merged.\newline

@@

In a last post processing step \texttt{MSF} identifies the trigger
points of the modulated sub-graphs. These trigger genes are the
sources of the sub-graphs with only outgoing edges. These genes can be
interpreted as the possible entry points of perturbation from where
the stimulus causes downstream effects. In the same spirit the most
downstream genes of the modulated sub-graph are identified and defined
as sinks. Sinks can be interpreted as the effectors where the
integrated information within the signal transduction network is set
to action. Due to circular loops not all sub-graphs must have sources or
sinks.

@@

To assess the robustness and stability of our method, Poisson
distributed noise was added to the sample read counts \TODO{more details;what samples;what DGEA}. Then DGEA was
carried out on the disturbed data with the same parameters as for the
native data, followed by analysis with \texttt{MSF}. This procedure was carried out 100 times
and every time the genes from the sub-graphs identified from noisy
data were compared to the genes of sub-graphs identified from the
native data. This was also done for the DEG obtained for each run, a
cutoff was used to compare the retrieved DEG from noisy data to the
native data.  The robustness of \texttt{MSF} and the DEG analysis for
the three time-points are shown in~(Fig.~2) \TODO{which time points}. We were very stringent in
adding noise to the counts, which could be seen by the DEG analysis
robustness. For the last time-point (48h~p.i.) 84~\% genes of the
sub-graphs from native data were retrieved, for 24~h~p.i. robustness was
about 79~\%. Robustness was 71~\% at 6~h~p.i., this could be due to that the
infection merely started and the signal transdcution was weak as
compared to the later time-points.

Simulation testing was also conducted on \texttt{MSF}, the objective
was to find less sources than observed in the sub-graphs identified by
\texttt{MSF} from native data. Simulation results showed a chance of
less than 10 percent to get more sources than observed \TODO{we agreed on massively expanding this section or skip it alltogether}.

@@

To demonstrate its usefulness \texttt{MSF} is applied to RNAseq data
set of primary human monocyte-derived macrophages (MDMs) infected with
Ebola virus (GSE84188)~\cite{Olejnik}. Ebola Virus (EBOV) belongs to
the Filoviridea family; filamentous, enveloped and single stranded RNA
viruses. EBOV causes hemorrhagic fever in humans, inducing the host
innate and adaptive immune response to be unable to control virus
infection~\cite{Prins}. Until now there are no approved antiviral
drugs for the treatment of Ebola virus infection~\cite{Konde,Rhein}.
The initial targets of EBOV are the macrophages and dendritic immune
cells~\cite{Falasca,Rhein}. EBOV inhibits the critical innate immune
response of the host, which includes the activation of alpha/beta
interferon (IFN-$\alpha / \beta$)~\cite{Prins,Konde,Cardenas}. The
Ebola viral protein VP35 targets the host type~I interferon
IFN-$\alpha / \beta$ to block the early antiviral immune response
~\cite{Prins,Konde,Falasca,Cardenas,Olejnik}. It has been proposed
that IFN-$\alpha / \beta$ should be tested against Ebola for its
antiviral activity through clinical trials~\cite{Konde}. The aim of
testing the EBOV infection data with \texttt{MSF} was to identify the
modulated sub-graphs and check if \texttt{MSF} is able to identify the
IFN-$\alpha / \beta$ gene as one of the sources. \TODO{why this data set?}

EBOV infection count data was downloaded from GEO
(GSE84188). Differential gene expression analysis was performed on the
count data with edgeR package (version 3.4.2)~\cite{edgeR}. The DEG
list generated by edgeR was used as an input for \texttt{MSF} \TODO{above you said MSF does not use a DEG gene list}. The
EBOV infection experiments describe the course of infection at three time-points 6, 24 and 48 hour post
infection (h~p.i.). For the earliest time point at 6 h~p.i., five modulated sub-graphs were
identified \TODO{with how many genes each?}. Most of the genes part of the sub-graphs were chemokines
(CXCL10, CCL8) and Interleukin genes (IL6, IL27, IL23) which serve as
a subset of CD4+ T helper cells. IFNB1 and IFNA1 were both identified
as two of the possible sources. Most of the sources identified by
\texttt{MSF} were type~I interferon induced genes. At 24h~p.i. nine
modulated sub-graphs were identified, IFNA1 was identified as one of
the sources. For the last time-point again IFNB1 and IFNA1 were
identified as the two sources out of the possible sources from seven
modulated sub-graphs identified.

\TODO{bis da!!}

@@

\subsection*{Modulated sub-graphs at 6h~p.i.}

 Three main modulated sub-graphs identified by \texttt{MSF} at 6h~p.i.
 are shown in~(Fig.~3). The graph shows different colored nodes as
 genes associated to KEGG pathways and how they cross talk. The KEGG
 association and visualization on the modulated sub-graphs was done
 using StringApp~\cite{StringApp} in Cytoscape~\cite{Cyto}. The graph
 unveils how more than one stimulus coincides to prevent the
 progression of infection. Sub-graph 1 shows toll-like receptor,
 cytokine, chemokine and jak-stat genes leading to TNF and eventually
 going into apoptosis. Sub-graph 2 shows Extra-cellular matrix (ECM)
 receptor interaction genes which are suggested to be in interaction
 with Ebola glycoprotein (GP)~\cite{Veljkovic}. ECM receptor, Focal
 adhesion and P13k-Akt signaling genes are the sources from one end of
 the sub-graph. P13k-Akt signaling goes to FoxO signaling which
 eventually leads to cell cycle, which at later time-points is
 disrupted. From the other end of sub-graph 2 toll-like recpetor
 combined with cytosolic DNA sensing and cytokine-chemokine signaling
 activates Jak-stat which leads to the start of apoptosis with a
 activation of a few apoptosis genes. Cytokine is also seen to sink
 into natural killer mediated cytotoxicity. Sub-graph 3 shows how some
 of the cytokine and chemokine genes activate the natural killer
 mediated cytotoxicity and apoptosis.

@@

At 6h~p.i. Cytokegg identified the sub-graph genes to be enriched in
Toll-like receptor signaling, TNF signaling, IL-17 signaling and
NF-kappa signaling pathways, as also stated by~\cite{Olejnik}. The
earliest response EBOV induces on cytokine is via TLR4-mediated
signaling~\cite{Olejnik}, Gene enrichment analysis of sub-graph genes
at 6h~p.i. showed toll-like receptor signaling being most significantly
dis-regulated pathway, this important signaling pathway was not
identified by SPIA along with RIG-I-like receptor signaling pathway,
TNF signaling pathway and IL-17 signaling (Table 1 in the supplement
material). On the later time-points \texttt{MSF} and SPIA showed
consensus on the important pathways e.g Chemokine signaling,
Cytokine-Cytokine receptor interaction, NF-kappa B signaling,
Cytosolic DNA-sensing and RIG-I-like receptor signaling as being
dis-regulated (Table 2-3 in the supplement).

@@

@@

\begin{figure*}
	\centering
	\fbox{
		\begin{minipage}{17 cm}
	\includegraphics[width=1.05\linewidth]{SeperatedMergedNetwork.png}
	\caption{Shows the three modulated sub-graphs identified by
          MSF at 6h~p.i., where nodes are genes and edges are the
          interactions. The node coloring is associated to KEGG
          pathways referring to the colors in the legend. The graph
          edges are from Reactome. Important genes to EBOV infection
          as from literature are enlarged in the graph. The signal
          flow corresponding to each sub-graph is depicted as simple
          flow chart against each sub-graph.}
		\end{minipage}
}
\end{figure*}
